NLIZE: A Perturbation-Driven Visual Interrogation Tool for Analyzing and Interpreting Natural Language Inference Models

### 2018-03-28 paper/background.tex
Pass through first two sections

At the first glance, the task of natural language inference may seem less practical compared to other well-known NLP challenges such as machine translation; however, the ability to distinguish entailment and contradiction relationship is fundamental to understanding natural language at large.
%
Considering the ambiguousness of natural language and polysemy words, the inference task can become quite challenging (especially from learning algorithm's point of view). Take the following sentences as an example (here \textbf{P} refers to premise, \textbf{H} refers to hypothesis):  (\textbf{P}) Facebook's IPO electrified the general public; (\textbf{H1}) Facebook went public; (\textbf{H2}) General Electric went public; (\textbf{H3}) People ignored Facebook's IPO. The literal similarity between ``Electric'' and ``electrified'' may trick a model to predict \textbf{H2} as \emph{entailment}. A model likely will also fail to understand the link between ``went public'' and ``IPO'', therefore, mistake \textbf{H1} as \emph{neutral}.
%
%The grand challenge in Natural Language Processing (NLP) is to have machine to acquire
%deep comprehension of textual information. Major tasks include Natural Language Inference,
%Machine Translation, Questions Answering, Summarization, and etc.
%Recently, neural networks models have performed strongly on these tasks.
%
Recently, Bowman et al. introduced a large corpus~\cite{BowmanAngeliPotts2015} for the NLI tasks, which has helped spawn a new wave of effective neural network models for those tasks. In this work, we focus on the analysis of the decomposable attention model~\cite{parikh2016emnlp} on this dataset.

@@

\subsection{Neural Network Model in NLP}
%\shusen{What is the basic understanding of neural network What is the end-to-end}
There are many connections and distinctions between neural network models employed for vision and NLP tasks.
On the one hand, as discussed in the introduction, the discrete nature of words and sentences presents additional challenges for
interpreting the NLP model.
%
On the other hand, a majority of recent NLP neural networks share the nature of
end-to-end model, where the entire model operates as a black-box that takes
vectorized input and yields a final prediction for a specific task.
%

Many of the recent NLP end-to-end neural network models, despite having potentially drastically different network architectures, share a similar high-level design that consists of three distinct stages (encode, align, classify, see Fig.~\ref{fig:modelPipeline}).
%
The existence of shared conceptual structures means the proposed tool, although designed for NLI task, can be generalized for a much broader set of applications.

@@

These end-to-end models usually take pre-trained word vectors (numerical vector representation of individual words, in which the semantic similarities are expressed by distances in the high-dimensional vector space~\cite{MikolovSutskeverChen2013, PenningtonSocherManning2014}) as input, and then in the encoding stage, the pre-trained vectors are adjusted to the specialized task at hand. Subsequently, the next stage aims at finding alignment between words in the input sequences. For the NLI task, this means finding the corresponds between words from the premise sentence to the hypothesis sentence (see details in Fig.~\ref{fig:attention} and Section~\ref{sec:attention}). Finally, in the last stage, the alignment information and the encoded vector representations are aggregated and then used as features for a classification network (last part of the end-to-end neural network).
%
Since all these three stages of the model are trained jointly in an end-to-end fashion, it is important to explore the interaction between intermediate representations and predictions to make sense of how predictions are made inside such a model.

\subsection{Attention Mechanism}
\label{sec:attention}
Among the three stages, the second stage often constitutes a crucial part of the model as it determines where the classifier will focus for generating a prediction. The operation to compute the alignment between words in the input is referred to as the attention mechanism. 
The introduction of the attention mechanism~\cite{bahdanau2014neural} allows
pairwise interaction between internal representations of words. 
This interaction can be naturally explained as a form of alignment that exposes an interpretable layer in end-to-end neural networks.
%
%As the only interpretable part of the internal state of the network, understanding how attention information 
Recently, attention has contributed to many strongly performed NLP models

~\cite{parikh2016emnlp},~\cite{rush2015neural},~\cite{yang2016hierarchical},
~\cite{seo2016bidirectional},~\cite{schwartz2017high}.

@@

>>>>

At the first glance, the task of natural language inference may seem less practical compared to other well-known NLP challenges such as machine translation; however, the ability to distinguish entailment and contradiction relationship is fundamental to understanding natural language at large.
%
Considering the ambiguousness of natural language and polysemy words, the inference task can become quite challenging (especially from the learning algorithm's point of view). Take the following sentences as an example (here \textbf{P} refers to premise, \textbf{H} refers to hypothesis):  (\textbf{P}) Facebook's IPO electrified the general public; (\textbf{H1}) Facebook went public; (\textbf{H2}) General Electric went public; (\textbf{H3}) People ignored Facebook's IPO. The literal similarity between ``Electric'' and ``electrified'' may trick a model to predict \textbf{H2} as \emph{entailment}. A model likely will also fail to understand the link between ``went public'' and ``IPO'', therefore, mistake \textbf{H1} as \emph{neutral}.
%
%The grand challenge in Natural Language Processing (NLP) is to have machine to acquire
%deep comprehension of textual information. Major tasks include Natural Language Inference,
%Machine Translation, Questions Answering, Summarization, and etc.
%Recently, neural networks models have performed strongly on these tasks.
%
Recently, Bowman et al. introduced a large corpus~\cite{BowmanAngeliPotts2015} for NLI tasks, which has helped spawn a new wave of effective neural network models for those tasks. Here, we focus on the analysis of the decomposable attention model~\cite{parikh2016emnlp} on this dataset.

@@

\subsection{Neural Network Model in NLP}
%\shusen{What is the basic understanding of neural network What is the end-to-end}
There are many connections and distinctions between neural network models employed for vision and NLP tasks.
On the one hand, as discussed in the introduction, the discrete nature of words and sentences presents additional challenges for
interpreting the NLP model.
%
On the other hand, a majority of recent NLP neural networks share the nature of
end-to-end models, where the entire model operates as a black-box that takes
vectorized input and yields a final prediction for a specific task.
%

Many of the recent NLP end-to-end neural network models, despite having potentially drastically different network architectures, share a similar high-level design that consists of three distinct stages (encode, align, classify, see Fig.~\ref{fig:modelPipeline}).
%
The existence of shared conceptual structures means the proposed tool, although designed for NLI tasks, can be readily generalized for a much broader set of applications.

@@

These end-to-end models usually take pre-trained word vectors (numerical vector representation of individual words, in which the semantic similarities are expressed by distances in the high-dimensional vector space~\cite{MikolovSutskeverChen2013, PenningtonSocherManning2014}) as input, and then in the encoding stage, the pre-trained vectors are adjusted to the specialized task at hand. Subsequently, the next stage aims at finding alignment between words in the input sequences. For the NLI task, this means finding the correspondance between words from the premise sentence to the hypothesis sentence (see details in Fig.~\ref{fig:attention} and Section~\ref{sec:attention}). Finally, in the last stage, the alignment information and the encoded vector representations are aggregated and then used as features for a classification network (last part of the end-to-end neural network).
%
Since all these three stages of the model are trained jointly in an end-to-end fashion, it is important to explore the interaction between intermediate representations and predictions to make sense of how predictions are made inside such a model.

\subsection{Attention Mechanism}
\label{sec:attention}
Among the three stages, the second stage often constitutes a crucial part of the model as it determines where the classifier will focus for generating a prediction. The operation to compute the alignment between words in the input is referred to as the attention mechanism. 
The introduction of the attention mechanism~\cite{bahdanau2014neural} allows
pairwise interaction between internal representations of words. 
This interaction can be naturally explained as a form of alignment that exposes an interpretable layer in end-to-end neural networks.
%
%As the only interpretable part of the internal state of the network, understanding how attention information 
Recently, attention has contributed to many strongly performed NLP models~\cite{parikh2016emnlp,rush2015neural,yang2016hierarchical,seo2016bidirectional,schwartz2017high}.

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "paper_entailVis"
%%% End:

### 2018-03-28 paper/intro.tex
update task
Pass through first two sections

% (What is the high level motivation)
% What is the obstacle for effectively apply machine learning model
With the recent advances in neural network model, machine learning has gained unprecedented popularity, and its adaption has permeated many fields of studies.
%
However, from researchers to practitioners, one often need to overcome many obstacles during the training, debugging, and tuning processes to unleash the full potential of these models.
%
Interpreting the internal mechanism and analyzing how predictions are made are critical to both the design and deployment of a model.
More importantly, the ability to pinpoint where and how an error is made and come up with hypotheses for the cause of the error is the key to identifying the limitation and improving upon existing models.
However, providing meaningful answers to these questions is incredibly challenging and often regarded as impossible by many.

Recently, sizable research has been developed to combat the model interpretability challenges. Most notably, the works~\cite{SimonyanVedaldiZisserman2013, ZeilerFergus2014, YosinskiCluneNguyen2015, OlahMordvintsevSchubert2017, LiuShiLi2017, OlahSatyanarayanJohnson2018, BilalJourablooYe2018} on interpreting the convolution neural network (CNN) model from both the machine learning and visualization communities have revealed the enormous potential and immediate benefits of model interpretation. At the same time, these works also remind us how little we truly understand about the inner mechanism of these deep neural networks.

Compared to image classification task, natural language processing (NLP) systems have innate restrictions (e.g., the discrete nature of words) that make the road to interpretation more treacherous. For example, \emph{feature visualization}~\cite{OlahMordvintsevSchubert2017}, an often deployed technique for illustrating what image features a given part of the network (e.g., neuron, layer, or channel) captures, cannot be readily generalized to natural language. An image (pixel values) represents a continuous solution space, in which human can quickly recognize the existence (or the absence) of patterns or shapes. For natural language, even though we can encode a word as a vector~\cite{MikolovSutskeverChen2013, PenningtonSocherManning2014}, the word embedding space is still discretely defined, in which the interpolation of two vectors (i.e., two words) does not hold clear meaning.
%
These restrictions call for new avenues for solving the interpretation challenges of natural language models, which motivates the proposed work.

% Beside explore the internal states of the model, understand how different parts of the model work together and how one may affect one another is as or if not more important.
Moreover, despite many recent advances, the majority of the existing techniques study the model as an invariant object, where the model's behaviors are recorded and analyzed in an offline fashion.
%
However, the exploratory nature of the model interpretation analysis often leads to many \emph{``what if ...''} type of questions. Such as, what if we perturbed the current input? Will the prediction be stable? What if we change one of the critical internal states of the model? How would the modification affect the prediction? What if the current prediction is wrong? How and where could we apply minimal change to the model to produce the correct result? And how would it affect the internal state we care about? These types of queries form a natural way to gain understanding and come up with hypotheses of the model mechanisms by \emph{interrogating} how the components of a model interact with each other in a dynamic setting. Hypothetically, we can code specialized experiments for each of these scenarios. However, such a process is not only tedious but also ignores the iterative nature of the exploratory analysis. Often, these questions are not pre-determined. Instead, new exploration paths arise as we investigate and analyze previous observations.

In this work, we aim to provide immediate and informative answers to these \emph{``what if''} questions by combining the expressive power of visualization and direct online query/optimization of the neural network model. Instead of viewing the models as invariant objects, we approach the interpretability challenge by studying them in a dynamic environment. By employing a perturbation-driven paradigm, we probe the internal states of the model and examine how changes in one part of the pipeline (the input, internal states, and output prediction, see Fig.~\ref{fig:modelPipeline}) affect others, which in turn provides a new perspective to address the model interpretation problem.

We implement our vision of the perturbation-driven exploration in an interactive visualization system for natural language inference models~\cite{Parikh2016}.
However, the components of the visualization and the overall paradigm can be readily extended to other NLP tasks, such as question and answer, and text summarization, etc.
%
In the simplest form, the inference task asks the question about whether we can infer the content of \textbf{sentence B} from \textbf{sentence A} (\emph{entail}), or if they are contradict with each other (\emph{contradiction}), or whether two sentences talk about different/unrelated things (\emph{neutral}). Language inference addresses the fundamental challenge of measuring the semantic similarity of sentences, and is one the core NLP tasks (see detail in Section~\ref{sec:languageInference}).

One recent advance in neural natural language process models is the introduction of \emph{attention mechanisms}~\cite{VaswaniShazeerParmar2017} (Section~\ref{sec:attention}). Intuitively, the attention helps capture which part of the input the model is deemed more important for making a prediction. The attention is often represented as weights for individual words or pairs of words (i.e., the alignment between words in different sentences).
%
%\shusen{seems repeated some of the previous points}
There have been many speculations on how attention works in various models. By utilizing the proposed tool, the domain experts study how changes in sentence input, attention, or prediction affect each other, which helps them develop deeper intuitive and alternative hypotheses. 
%we can perturb the attention and observer the change in the prediction, perturb the sentence input and observe the effects on the attention. And more importantly, we also allow the user to change prediction, where we apply a constrained optimization to find the least amount of parameter update to produce the user assign label. 
%Another word, 
%Such type of exploration is not only essential to interpret the language inference problem, but can also be generalized to other NLP model utilized the attention mechanisms.
Moreover, we enhanced the standard visual encoding (e.g., as a bipartite graph or as a matrix) of the attention matrix by overlaying sentence linguistic structure to allow grammar-guided simplification of the visual representation.
%
Finally, as discussed in Section~\ref{sec:caseStudy}, the ability to examine how attention corresponds to the grammatical structure also enables domain experts to speculate about the potential benefits of including the linguistic structure in the design of the attention component of the model.

In summary, the key contributions of the proposed works are the following:
\begin{itemize}
  % \item Distill the visualization tasks for making sense of the role of attention in natural language inference (section~\ref{sec:task});
    \item Distill the perturbation-driven exploration paradigm from close examination of how domain experts conduct exploratory analysis on end-to-end NLP neural network models;

    \item Present the NLIZE system that enriches such a paradigm by providing an intuitive environment where the domain experts can readily express experiment ideas and obtain instantaneous feedbacks; 
    %to  experiment with the intricate relationships among input sentences, model internals, and the prediction;
    %\item Perform error analysis by examining the stability of the predictions under perturbation;

    \item Introduce an optimization scheme for correcting a failed prediction based on the natural extension of the margin-infused relaxed algorithm (MIRA) to the neural network.
    % to minimally update the model to fix a failed prediction, which help reason about the  better understand the re the question about how t

    \item Enhance the visual encoding of the attention by imposing sentence linguistic structure to allow grammar-guided sentence simplification.
\end{itemize}
%
% In addition,

>>>>

% (What is the high level motivation)
% What is the obstacle for effectively apply machine learning model
Based on some stunning recent advances, machine learning and neural networks base deep learning approaches in particular have gained unprecedented popularity, and have been adopted in a wide variety of scientific applications.
%
However, researchers as well as practitioners often need to overcome many obstacles during the training, debugging, and tuning processes to realize the full potential of these models.
%
Interpreting the internal mechanism and analyzing how predictions are made is critical to both the design and deployment of a model.
More importantly, the ability to pinpoint where and how an error is made and propose hypotheses for the cause of the error is key to identifying the limitations of and improving upon existing models.
However, providing meaningful answers to these questions is challenging and has been described as impossible by many.

Recently, significant research has been developed to combat the model interpretability challenges.
Both the machine learning as well as the visualization community have proposed a number of promising techniques aimed at interpreting convolution neural networks (CNN)~\cite{SimonyanVedaldiZisserman2013, ZeilerFergus2014, YosinskiCluneNguyen2015, OlahMordvintsevSchubert2017, LiuShiLi2017, OlahSatyanarayanJohnson2018, BilalJourablooYe2018}. 
At the same time, these approaches also remind us how little we truly understand about the inner mechanism of such deep neural networks.

Compared to image classification tasks, natural language processing (NLP) systems have innate restrictions (e.g., the discrete nature of words) that make the road to interpretation even more challenging.
For example, \emph{feature visualization}~\cite{OlahMordvintsevSchubert2017}, an often deployed technique for illustrating what image features a given part of the network (e.g., neuron, layer, or channel) captures, cannot be readily generalized to natural language. 
An image (pixel values) represents a continuous solution space, in which human can quickly recognize the existence (or the absence) of patterns or shapes. 
For natural language, even though we can encode a word as a vector~\cite{MikolovSutskeverChen2013, PenningtonSocherManning2014}, the word embedding space is still discretely defined, in which the interpolation of two vectors (i.e., two words) does not hold clear meaning.
%
These restrictions call for new avenues for solving the interpretation challenges of natural language models, which motivates the proposed work.

% Beside explore the internal states of the model, understand how different parts of the model work together and how one may affect one another is as or if not more important.
Despite a number of recent advances, the majority of existing techniques study the model as an invariant object, where the model's behaviors are recorded and analyzed in an offline fashion.
%
However, the exploratory nature of the model interpretation often leads to many \emph{``what if ...''} type of questions. 
Such as, what if we perturbed the current input?
Will the prediction be stable?
What if we change one of the critical internal states of the model? 
How would the modification affect the prediction? 
What if the current prediction is wrong? 
How and where could we apply minimal change to the model to produce the correct result? And how would it affect the internal state we care about? These types of queries form a natural way to gain understanding and develop hypotheses of the model mechanisms by \emph{interrogating} how the components of a model interact with each other in a dynamic setting.
Hypothetically, we can code specialized experiments for each of these scenarios.
However, such a process is not only tedious but also ignores the iterative nature of the exploratory analysis. Often, these questions are not pre-determined.
Instead, new exploration paths arise as we investigate and analyze previous observations.

Here, we aim to provide immediate and informative answers to these \emph{``what if''} questions by combining the expressive power of visualization and direct online query/optimization of the neural network model. Instead of viewing the models as invariant objects, we approach the interpretability challenge by studying them in a dynamic environment. By employing a perturbation-driven paradigm, we probe the internal states of the model and examine how changes in one part of the pipeline (the input, internal states, and output prediction, see Fig.~\ref{fig:modelPipeline}) affect others, which in turn provides a new perspective to address the model interpretation problem.

We implement our vision of the perturbation-driven exploration in an interactive visualization system for natural language inference models~\cite{Parikh2016}.
However, the components of the visualization and the overall paradigm can be readily extended to other NLP tasks, such as question and answer, text summarization, etc.
%
In the simplest form, the inference task asks the question about whether \textbf{sentence A} and \textbf{sentence B}: (1) are \emph{entailed} (one can infer \textbf{B} from \textbf{A}; (2) \emph{contradict} (\textbf{B} disagrees with \textbf{A}) or (3) are \emph{neutral} (\textbf{A} and \textbf{B} talk about different/unrelated thins
% we can infer the content of \textbf{sentence B} from \textbf{sentence A} (\emph{entail}), if they contradict with each other (\emph{contradiction}), or whether two sentences talk about different/unrelated things (\emph{neutral}). 
Language inference addresses the fundamental challenge of measuring the semantic similarity of sentences, and is one the core NLP tasks (see Section~\ref{sec:languageInference} for details).

One recent advance in neural natural language process models is the introduction of \emph{attention mechanisms}~\cite{VaswaniShazeerParmar2017} (Section~\ref{sec:attention}). Intuitively, the attention helps capture which part of the input the model is deemed more important for making a prediction. The attention is often represented as weights for individual words or pairs of words (i.e., the alignment between words in different sentences).
%
%\shusen{seems repeated some of the previous points}
There have been many speculations on how attention works in various models. By utilizing the proposed tool, the domain experts study how changes in sentence input, attention, or prediction affect each other, which helps them develop deeper intuitive and alternative hypotheses. 
%we can perturb the attention and observer the change in the prediction, perturb the sentence input and observe the effects on the attention. And more importantly, we also allow the user to change prediction, where we apply a constrained optimization to find the least amount of parameter update to produce the user assign label. 
%Another word, 
%Such type of exploration is not only essential to interpret the language inference problem, but can also be generalized to other NLP model utilized the attention mechanisms.
Moreover, we propose to enhance the standard visual encoding (e.g., as a bipartite graph or as a matrix) of the attention matrix by overlaying sentence linguistic structure to allow grammar-guided simplification of the visual representation.
%
Finally, as discussed in Section~\ref{sec:caseStudy}, the ability to examine how attention corresponds to the grammatical structure also enables domain experts to speculate about the potential benefits of including the linguistic structure in the design of the attention component of the model.

In summary, the key contributions of the proposed works are the following:
\begin{itemize}
  % \item Distill the visualization tasks for making sense of the role of attention in natural language inference (section~\ref{sec:task});
    \item A perturbation-driven exploration paradigm derived from close examination of how domain experts conduct exploratory analysis on end-to-end NLP neural network models;

    \item The NLIZE system that enriches such a paradigm by providing an intuitive environment that allows domain experts to readily express hypotheses and obtain instantaneous feedbacks; 
    %to  experiment with the intricate relationships among input sentences, model internals, and the prediction;
    %\item Perform error analysis by examining the stability of the predictions under perturbation;

    \item An optimization scheme for correcting a failed prediction based on the natural extension of the margin-infused relaxed algorithm (MIRA) to the neural network; and
    % to minimally update the model to fix a failed prediction, which help reason about the  better understand the re the question about how t

    \item A visual encoding of the attention by imposing sentence linguistic structure to allow grammar-guided sentence simplification.
\end{itemize}
%
% In addition,

### 2018-03-28 paper/paper_entailVis.tex
Pass through first two sections

%% In preprint mode you may define your own headline.
%\preprinttext{To appear in IEEE Transactions on Visualization and Computer Graphics.}
\usepackage{amsmath}
\usepackage[pdftex,dvipsnames]{xcolor}  % Coloured text etc.

\newcommand{\zhimin}[1]{\textcolor{magenta}{[ZHIMIN: #1]}}
\newcommand{\shusen}[1]{\textcolor{red}{[#1]}}

@@

%% Abstract section.
% What are we doing in this paper
\abstract{
With the recent advances in deep learning, neural network models have obtained state-of-the-art performances for many linguistic tasks in natural language processing. The disruptive progress also brings enormous challenges. The opaque nature of neural network model leads to hard-to-debug-systems and difficulty interpreting mechanisms. In this work, we introduce a visualization system that through a tight yet flexible integration between visualization elements and the underlying model allows a user to interrogate the model by perturbing the input, the internal state, and prediction while observing changes in other parts of the pipeline. We use the natural language inference problem as an example to illustrate how a perturbation-driven paradigm can help domain experts assess the potential limitation of a model, probe its inner states and interpreting and hypothesizing fundamental model mechanisms such as attention.
} % end of abstract

>>>>

%% In preprint mode you may define your own headline.
%\preprinttext{To appear in IEEE Transactions on Visualization and Computer Graphics.}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[pdftex,dvipsnames]{xcolor}  % Coloured text etc.

\newcommand{\zhimin}[1]{\textcolor{magenta}{[ZHIMIN: #1]}}
\newcommand{\shusen}[1]{\textcolor{red}{[#1]}}
\newcommand{\taoli}[1]{\textcolor{orange}{[#1]}}

@@

%% Abstract section.
% What are we doing in this paper
\abstract{
With the recent advances in deep learning, neural network models have obtained state-of-the-art performances for many linguistic tasks in natural language processing. However, this rapid progress also brings enormous challenges. The opaque nature of neural network model leads to hard-to-debug-systems and difficult to interpret mechanisms. Here, we introduce a visualization system that through a tight yet flexible integration between visualization elements and the underlying model allows a user to interrogate the model by perturbing the input, the internal state, and prediction while observing changes in other parts of the pipeline. We use the natural language inference problem as an example to illustrate how a perturbation-driven paradigm can help domain experts assess the potential limitation of a model, probe its inner states and interpret and hypothesize about fundamental model mechanisms such as attention.
} % end of abstract

### 2018-03-28 paper/relatedWorks.tex
Pass through first two sections

\section{Related Works}
With the increasing demands for model accountability (e.g., what is the evidence for making the prediction) and model fairness (e.g., is the prediction affected by the bias in the training data),
the interpretability of the machine learning model held significant implications. As a result, a wealth of research has been developed that focused on the explanation and interpretation of neural network models from both machine learning and visualization community.

>>>>

\section{Related Work}
With the increasing demands for model accountability (e.g., what is the evidence for making the prediction) and model fairness (e.g., is the prediction affected by the bias in the training data),
the interpretability of the machine learning model held significant implications. As a result, a wealth of research has been developed that focused on the explanation and interpretation of neural network models from both machine learning and visualization community.
